Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-03-31 joint_inf.tex
writing out explicit steps

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
%EM Not sure how tau induces an ordered set of tip labels-- are you saying that the tree defines the labels or something about its structure?
Following Section 8.6 of \cite{Semple2003-em}, $\tau$ induces an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ across sites.
For the binary symmetric model, a realization of $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet\triangleq\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
Indeed, we can start by including those tips in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$. Then, because the probability of observing a particular site split is equivalent to the probability of its complement under the binary symmetric model, we can take the complement if needed to exclude tip $\nSiteRows$ from $\siteSplit$.
Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ where $\mathcal{P}$ denotes the powerset.
%EM inducing again.
The topology $\tau$ also induces an ordered set of internal nodes $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of characters $\ancestralSplit\subseteq\ancestralSplitSet\triangleq\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
The probability of observing an ancestral state split, conditional on a site split, will not be invariant to taking its complement, and thus the need for enumerating the entire set of internal nodes.
The number of unique ancestral state splits is $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$.

With this split formulation, define the mappings from patterns to splits as described above
$$
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet), \ \ancestralToSplit:\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
% the following seemed not necessary and a little confusing (what are j and k?)
% so that, for
% $$
% \patternToSplit(\alignmentColumn_i)=\siteSplit_j, \ \ancestralToSplit(\ancestralStateColumn_i)=\ancestralSplit_k.
% $$
Given a site-pattern-valued random variable $H$ and an ancestral-state-valued random variable $H$, define random variables
$$
\siteSplitRV := \patternToSplit(\alignmentColumnRV), \ \ancestralSplitRV := \ancestralToSplit(\ancestralStateColumnRV).
$$
We can now write the $i$th factor of \eqref{eq:full_likelihood} as
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)=P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)
$$
for some $j$ and $k$.
Taking the maximum of the probability of observing a particular column $i$ yields
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &=
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
&= P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
%EM Just noticed that eta doesn't have to be a real partition, because they don't have to be disjoint, right?
Given $(\tau, t)$ there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

\begin{align}
L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\
%EM I guess the ith component of H is h_i?
                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
%EM Indexing broken here: we are taking a product over i, but i doesn't appear in the expression.
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
%EM As I fixed above and described in person, \ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) is not what you want here, because it will inflate the value by |\ancestralSplitPartition_j(\tau, t)| . This also appears below.
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ appears.

@@

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
%EM I dropped the \ancestralSplitPartition(\siteSplit, \tau, t) notation above. You can revert it if you like, but I prefer the bold eta notation as an ordered list of sets.
A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; define $\ancestralSplitPartition(\siteSplit, \tau, t)$ as the partition of $\mathcal{P}(\ancestralSplitSet)$ into those most likely ancestral splits for that particular site split, topology and set of branch lengths.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
The probabilities of observing ancestral splits for site split $\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
%EM Seems to me that this just requires one character change so would be p(1-p)^4.
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^2(1-p)^3,
$$
$$
%EM Have you told us in this example what the labeled topology is?
%Erick stopped here.
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
Here, the partition into most likely ancestral states will proceed as
$$
P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p) \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5
$$
where the partition is the single element $\ancestralSplitPartition_j(\tau, p)=\emptyset$.
For the site split $\{1,3\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^3(1-p)^2,
$$
where
$$
P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p) \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3
$$
with the most likely partition $\ancestralSplitPartition_j(\tau, p)=\{\emptyset,\{1,2\}\}.$

@@

Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
Define, e.g., $p_0 = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})$.
Let
$$
    a_{0} = p_{13},
$$
$$
    a_{1} = p_1+p_2+p_{12}+p_{23},
$$
$$
    a_{2} = p_3+p_{12}+p_{23}+p_{123},
$$
$$
    a_{3} = p_1+p_3+p_{12}+p_{23},
$$
$$
    a_{4} = p_2+p_{12}+p_{23}+p_{123}.
$$
Then the full likelihood is bounded above by one of three functions:
\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\}) \\
    &\qquad \le (3-a_{1})\log(1+x')+(a_{1}+a_{3})\log(1-x')\\
    &\qquad \qquad +(3-a_{2})\log(1+y')+(a_{2}+a_{4})\log(1-y')\\
    &\qquad \qquad +(1-a_{0})\log(1+w')+a_{0}\log(1-w')+p_{0}\log(4)\\
    &\qquad \le (3-a_{1})\log\frac{2(3-a_{1})}{3+a_{3}}+(a_{1}+a_{3})\log\frac{2(a_{1}+a_{3})}{3+a_{3}}\\
    &\qquad \qquad +(3-a_{2})\log\frac{2(3-a_{2})}{3+a_{4}}+(a_{2}+a_{4})\log\frac{2(a_{2}+a_{4})}{3+a_{4}}\\
    &\qquad \qquad +(1-a_{0})\log(1+w')+a_{0}\log(1-w')+p_{0}\log(4)\\
    &\qquad \triangleq L^{p,1}_{\tau_1,\tau_1}(\mathbf{a}_{xy});
\end{align*}
\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\}) \\
    &\qquad \le (3-a_{1}-2a_{0})\log(1+x')+(a_{1}+a_{3}+2a_{0})\log(1-x')\\
    &\qquad \qquad +(3-a_{2})\log(1+y')+(a_{2}+a_{4})\log(1-y')+\log(1+w')+p_{0}\log(4)\\
    &\qquad \le (3-a_{1}-2a_{0})\log\frac{2(3-a_{1}-2a_{0})}{3+a_{3}}+(a_{1}+a_{3}+2a_{0})\log\frac{2(a_{1}+a_{3}+2a_{0})}{3+a_{3}}\\
    &\qquad \qquad +(3-a_{2})\log\frac{2(3-a_{2})}{3+a_{4}}+(a_{2}+a_{4})\log\frac{2(a_{2}+a_{4})}{3+a_{4}}\\
    &\qquad \qquad +\log(2)+p_{0}\log(4)\\
    &\qquad \triangleq L^{p,2}_{\tau_1,\tau_1}(\mathbf{a}_{xy});
\end{align*}
\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\}) \\
    &\qquad \le (3-a_{1})\log(1+x')+(a_{1}+a_{3})\log(1-x')\\
    &\qquad \qquad +(3-a_{2}-2a_{0})\log(1+y')+(a_{2}+a_{4})+2a_{0}\log(1-y')+\log(1+w')+p_{0}\log(4)\\
    &\qquad \le (3-a_{1})\log\frac{2(3-a_{1})}{3+a_{3}}+(a_{1}+a_{3})\log\frac{2(a_{1}+a_{3})}{3+a_{3}}\\
    &\qquad \qquad +(3-a_{2}-2a_{0})\log\frac{2(3-a_{2}-2a_{0})}{3+a_{4}}+(a_{2}+a_{4}+2a_{0})\log\frac{2(a_{2}+a_{4}+2a_{0})}{3+a_{4}}\\
    &\qquad \qquad +\log(2)+p_{0}\log(4)\\
    &\qquad \triangleq L^{p,3}_{\tau_1,\tau_1}(\mathbf{a}_{xy}).
\end{align*}

>>>>

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
%EM Not sure how tau induces an ordered set of tip labels-- are you saying that the tree defines the labels or something about its structure?
Following Section 8.6 of \cite{Semple2003-em}, we define an ordered set of tip labels $\{1,\ldots,\nSiteRows\}$ on $\tau$ across sites.
For the binary symmetric model, a realization of $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
Indeed, we can start by including those tips in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$. Then, because the probability of observing a particular site split is equivalent to the probability of its complement under the binary symmetric model, we can take the complement if needed to exclude tip $\nSiteRows$ from $\siteSplit$.
Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ where $\mathcal{P}$ denotes the powerset.
%EM inducing again.
For topology $\tau$, we also define an ordered set of internal nodes $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ with a similar subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ describing a realization of $\ancestralStateColumn_i$.
The probability of observing an ancestral state split, conditional on a site split, will not be invariant to taking its complement, and thus the need for enumerating the entire set of internal nodes.
The number of unique ancestral state splits is $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$.

With this split formulation, define the mappings from patterns to splits as described above
$$
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet), \ \ancestralToSplit:\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
% the following seemed not necessary and a little confusing (what are j and k?)
% so that, for
% $$
% \patternToSplit(\alignmentColumn_i)=\siteSplit_j, \ \ancestralToSplit(\ancestralStateColumn_i)=\ancestralSplit_k.
% $$
Given a site-pattern-valued random variable $\alignmentColumnRV$ and an ancestral-state-valued random variable $\ancestralStateColumnRV$, define random variables
$$
\siteSplitRV := \patternToSplit(\alignmentColumnRV), \ \ancestralSplitRV := \ancestralToSplit(\ancestralStateColumnRV).
$$
We can now write the $i$th factor of \eqref{eq:full_likelihood} as
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)=P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)
$$
for some $j$ and $k$.
Taking the maximum of the probability of observing a particular column $i$ yields
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &=
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
&= P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
%EM Just noticed that eta doesn't have to be a real partition, because they don't have to be disjoint, right?
%das: I think they do have to be disjoint. The partition is into "most likely" and "not most likely," which is disjoint and exhausts the set. Am I missing something?
Given $(\tau, t)$ there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

\begin{align}
L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\
%EM I guess the ith component of H is h_i?
    %das: Not sure what the question/comment is here.
                             &= \max_{\fullAncestralStates} \ \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
%EM Indexing broken here: we are taking a product over i, but i doesn't appear in the expression.
    %das: This is intentional, though: it allows us to proceed to the following line, grouping factors by site pattern since their probabilities don't depend on $i$. I can try to think of how to make this more clear, but I thought it was a pretty key step to be able to pass to site patterns, and the reason for the machinery involved in defining ancestral state partitions.
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
%EM As I fixed above and described in person, \ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) is not what you want here, because it will inflate the value by |\ancestralSplitPartition_j(\tau, t)| . This also appears below.
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ appears.

@@

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
%EM I dropped the \ancestralSplitPartition(\siteSplit, \tau, t) notation above. You can revert it if you like, but I prefer the bold eta notation as an ordered list of sets.
A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; define $\ancestralSplitPartition(\siteSplit, \tau, t)$ as the partition of $\mathcal{P}(\ancestralSplitSet)$ into those most likely ancestral splits for that particular site split, topology and set of branch lengths.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
The probabilities of observing ancestral splits for site split $\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
%EM Seems to me that this just requires one character change so would be p(1-p)^4.
%das: I think it's wrong, but in the other direction: two character changes to get to the differing internal state, then one between the two internal states.
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
%EM Have you told us in this example what the labeled topology is?
%das: Nope! I wanted to be simple, but it seems maybe too simple. Maybe I should move the Felsenstein/Farris bit to this example?
%Erick stopped here.
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
Here, the partition into most likely ancestral states will proceed as
$$
P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p) \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5
$$
where the partition is the single element $\ancestralSplitPartition_j(\tau, p)=\emptyset$.
For the site split $\{1,3\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^3(1-p)^2,
$$
where
$$
P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p) \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3
$$
with the most likely partition $\ancestralSplitPartition_j(\tau, p)=\{\emptyset,\{1,2\}\}.$

@@

Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
For compactness, define, e.g., $p_0 = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})$, keeping in mind it is a function of parameters $x,y$.
The full Farris log likelihood can take one of three values---assume for now the we are only interested in the ancestral state split $\{2\}$ for the site split $\{1,3\}$.
We then write the log likelihood as
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
    &=        p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{1}  \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + p_{2}  \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + p_{3}  \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + p_{12} \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{23} \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + p_{123}\cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{2}  \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{3}  \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
    &\qquad + p_{12} \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
    &\qquad + p_{13} \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
    &\qquad + p_{23} \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
    &\qquad + p_{123}\cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
\end{align*}
Seeking an upper bound, we make use of the fact that, for $x',y'\in[0,1]$,
$$
(1-x'^2+y'^2-x'^2y'^2) = (1-x'^2)(1+y'^2) = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y),
$$
with similar substitutions for $(1+x'^2-y'^2-x'^2y'^2)$ and $(1-x'^2-y'^2+x'^2y'^2)$.
Doing so results in
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1+y')) \\
    &\qquad + p_{2}  \cdot\log((1+x')(1+y')(1-y')) \\
    &\qquad + p_{3}  \cdot\log((1+x')(1-x')(1+y')) \\
    &\qquad + p_{12} \cdot\log((1+x')(1-x')(1+y')(1-y')) \\
    &\qquad + p_{23} \cdot\log((1+x')(1-x')(1+y')(1-y')) \\
    &\qquad + p_{123}\cdot\log((1+x')(1+y')(1-y')) \\
    &\qquad + p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{2}  \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{3}  \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
    &\qquad + p_{12} \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
    &\qquad + p_{13} \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
    &\qquad + p_{23} \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \\
    &\qquad + p_{123}\cdot\log((1+x')^2   (1+w')(1+y')(1-y')).
\end{align*}
Grouping together like terms,
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + (1-p_{0}-p_{13})\cdot\log(1+x') \\
    &\qquad + (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
    &\qquad + (1-p_{0}-p_{13})\cdot\log(1+y') \\
    &\qquad + (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
    &\qquad + (2-p_{1}-p_{2}-p_{12}-p_{23})\cdot\log(1+x') \\
    &\qquad + (p_{1}+p_{2}+p_{12}+p_{23})\cdot\log(1-x') \\
    &\qquad + (2-p_{3}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
    &\qquad + (p_{3}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
    &\qquad + (1-p_{13})\cdot\log(1+w') \\
    &\qquad + p_{13}\cdot\log(1-w'),
\end{align*}
which equals
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + (3-p_{0}-p_{13}-p_{1}-p_{2}-p_{12}-p_{23})\cdot\log(1+x') \\
    &\qquad + (p_{2}+p_{3}+2(p_{1}+p_{3}+p_{12}+p_{23}))\cdot\log(1-x') \\
    &\qquad + (3-p_{0}-p_{13}-p_{3}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
    &\qquad + (p_{2}+p_{3}+2(p_{12}+p_{23}+p_{123}))\cdot\log(1-y') \\
    &\qquad + (1-p_{13})\cdot\log(1+w') \\
    &\qquad + p_{13}\cdot\log(1-w').
\end{align*}
To make matters further compact, let
$$
a_{1} = 3-p_{0}-p_{13}-p_{1}-p_{2}-p_{12}-p_{23},
$$
$$
a_{2} = p_{2}+p_{3}+2(p_{1}+p_{3}+p_{12}+p_{23}),
$$
$$
a_{3} = 3-p_{0}-p_{13}-p_{3}-p_{12}-p_{23}-p_{123},
$$
$$
a_{4} = p_{2}+p_{3}+2(p_{12}+p_{23}+p_{123}),
$$
where
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{13} \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + a_{1}\cdot\log(1+x') \\
    &\qquad + a_{2}\cdot\log(1-x') \\
    &\qquad + a_{3}\cdot\log(1+y') \\
    &\qquad + a_{4}\cdot\log(1-y') \\
    &\qquad + (1-p_{13})\cdot\log(1+w') \\
    &\qquad + p_{13}\cdot\log(1-w').
\end{align*}
Applying Gibbs' inequality to the first two summands, and maximizing over the remaining terms yields
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
    &\le      p_{0}  \cdot\log(p_{0}) \\ 
    &\qquad + p_{13} \cdot\log(p_{13}) \\
    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
    &\qquad + (1-p_{13})\cdot\log(2(1-p_{13})) \\
    &\qquad + p_{13}\cdot\log(2p_{13}).
    &\qquad := L^{p,1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}
